Natural Selection in the 20th Century

### 2020-11-09 why-negative-selection.Rmd
Add abstract. Rewrite intro. Rewrite conclusion. Reorganize and shorten. Move material to appendix.

---
title: "Natural Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "July 2020"
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
---

@@

Several gaps in our knowledge remain. First, while existing work examines
contemporary populations, we would like to extend the analysis over time, since
natural selection in the same direction over multiple generations
could lead to significant effects in the long run. Second, there are concerns
about ascertainment bias in the samples used. This could potentially bias
estimates of the magnitude of natural selection. Lastly, we know little about
the *mechanisms* behind natural selection in contemporary populations.

Here we attempt to answer these questions, using data from UK Biobank. We use
questions on respondents' number of children, and their number of siblings, to 
reconstruct two generations of family size. We then relate these to 33 different
polygenic scores for a variety of characteristics. This analysis provides a 
broader picture of natural selection in the UK than has previously been available.

Our key results are as follows. First, comparing natural selection across the
two generations, there are few signs of change in size or direction over time.
Most polygenic scores are selected in the same direction in both generations. We
also see little evidence of changes across birth cohorts within either
generation.

Second, natural selection is concentrated among population subgroups which are
under-represented in the sample, including less-educated people and poorer
households. When we weight the sample to correct for ascertainment bias, many
effect sizes are larger, sometimes much larger. We emphasize that our weights
are only a partial solution to ascertainment bias in the sample. As a result,
much uncertainty remains about the strength of natural selection in the UK
population, but we can be reasonably sure that it is larger than naïve estimates
suggest.

Lastly, patterns in natural selection give us clues to the underlying
mechanism. In particular, among mothers who were older at their first birth, and
among parents with fewer sexual partners, the direction of natural selection is
*reversed*: for example, among these subgroups, higher polygenic scores (PGS)
for educational attainment are associated with having more offspring, rather
than fewer. The effects of PGS can be decomposed into two channels: an effect on
age at first live birth, and an opposite-signed effect on number of children
conditional on age at first live birth.

This pattern can be explained by economic theories of fertility
[@becker1960economic]. In these, higher potential earnings have two opposite
effects on fertility: a fertility-increasing *income effect* (higher income
makes children more affordable), and a fertility-lowering *substitution effect*
(time spent on childrearing has a higher cost in foregone earnings). Among
couples who can take advantage of coordination in childcare activities, the
income effect dominates; among single parents, or those in unstable
relationships, the substitution effect dominates. Supporting this explanation,
polygenic scores' raw effects on fertility are negatively correlated with their
effects on income; on the other hand, controlling for age at first live birth,
scores' effects on fertility and on household income are *positively*
correlated. These two results are compatible with the substitution effect and the
income effect respectively.

@@

# Data

Data is taken from UK Biobank. Polygenic scores were normalized to mean 0,
variance 1.

@@

We first show raw changes in polygenic scores over birth years, within the entire UK Biobank
sample. Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores by 5-year
birth intervals. Several scores show consistent increases or declines over this 
30-year period, of the order of 5% of a standard deviation.

@@

To examine the effects of sampling, we start by measuring the strength of
natural selection within different subgroups. To do this we regress respondents' 
polygenic scores on their number of children ($y_i$):

\[
y_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
\]

@@

The effect size $\beta$ measures the strength of natural selection within the
sample or a given subgroup. In fact, since polygenic scores are normalized to
mean 0, $\beta$ is the expected polygenic score among children of the sample or
subgroup [@beauchamp2016genetic]. We begin with basic demographic variables
including education, income and sex. These are all potential sources of ascertainment
bias: as well as the ascertainment for income and education, mentioned above, the
sample sex ratio  skews `r female_ratio * 100`% female.

Figures \@ref(fig:plot-educ-level) and \@ref(fig:plot-income) plot effect sizes
for each polygenic score. Figure \@ref(fig:plot-educ-level) groups respondents
by age of completing full-time education. Figure \@ref(fig:plot-income) groups
respondents by household income. Effect sizes are larger and more significant
for the lowest income category, and for the lowest education category. Note that
the overall effect is not a simple average of the effect among the different
subgroups, because polygenic scores may also shift respondents between the
subgroups. For example, a high PGS for educational attainment may correlate
with fewer children among early school leavers, but may also increase the
age at which a respondent leaves school.

```{r plot-educ-level, fig.cap = "Effect sizes on number of children by age left full-time education"}

@@

```{r plot-income, fig.cap = "Effect sizes on number of children by household income"}

@@

Figure \@ref(fig:plot-sexes) shows effect sizes of PGS on number of children
separately for males and females. Differences are particularly large for
educational attainment, height and MDD. Several PGS for mental illness and
personality traits are more selected for (or less against) among women, including 
major depressive disorder (MDD), schizophrenia and neuroticism, while 
extraversion is more selected for among men. Waist circumference and waist-hip 
ratio are less selected for among women, and educational attainment is more
selected against, though the difference is only significant for EA3.

```{r plot-sexes, fig.cap = "Effect sizes on number of children by sex. Lines show significant differences"}

These results show that ascertainment within UK Biobank could bias estimates
of the size of natural selection towards zero. We now focus in on variables 
specifically related to marriage markets and raising children. Whatever processes 
give rise to natural selection, these variables are likely to be involved; at
the same time, they may also be ascertained within the sample.

Figure \@ref(fig:plot-n-partners) splits males and females by lifetime number of
sexual partners, at the median value of 3. For both sexes, effects are larger
and more significant among those with more than 3 partners. Results are
unchanged if we control for respondents' age (Figure
\@ref(fig:plot-n-partners-controlled) in the appendix). Figure
\@ref(fig:plot-with-partner) splits respondents up by whether they were living
with a spouse or partner at the time of interview. Again, effects are larger
among those not living with a spouse or partner.

```{r plot-n-partners, fig.cap = "Effect sizes on number of children by number of sexual partners", fig.height = 4}

```{r plot-with-partner, fig.cap = "Effect sizes on number of children by household type"}

res_with_partner %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner"),
        with_p    = as.character(grepl("TRUE", term)),
        Household = sprintf("%s (N = %s)", Household, 
                      prettyNum(hhtype_n[with_p], big.mark = ","))
      ) %>% 
      standard_ggplot(fill = Household, n_regs = n_regs, score_col = score_name)
```

@@

Our final variable is age at first live birth (AFLB), for females only. There is evidence
for genetic effects on AFLB [@Barban_2016]. There is also 
a close link between this variable and number of children born.

Figure \@ref(fig:plot-age-flb-cross) shows effect sizes estimated separately for
each tercile of AFLB. Several effects are strikingly
different across terciles. ADHD and MDD are selected for amongst the youngest
third of mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but is not
significantly selected among the youngest third. Similarly, several PGS of body
measurements are selected against only among older mothers. The correlation
between effect sizes for the youngest and oldest terciles is `r cor_age_flb`.
As before, recall that PGS may have two effects: they may change the number of
children within categories, but may also shift respondents between categories.

```{r plot-age-flb-cross, fig.cap = "Effect sizes on number of children, by age at first live birth terciles"}

@@

res_age_flb_cross %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(1)),
        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
        term       = add_n(term, "age_flb_cat")
      ) %>% 
      ggplot(aes(estimate, score_name, fill = term, 
        colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 3, direction = -1, 
          guide = guide_legend(title = "Age at first live birth")) +
        my_labs() +
        my_theme()

@@

To investigate this further, we regress number of children on PGS *controlling*
for AFLB. Figure \@ref(fig:plot-age-flb) shows the results, along with raw
effect sizes for comparison. Controlling for AFLB greatly reduces effect sizes.
This could be because the controlled regressions exclude childless women.
However, in `r opp_signed` out of `r length(score_names)` cases, effects
actually change sign when controls are added. The correlation between effect
sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`. So,
natural selection "effects" seem to come through two channels: an effect on
AFLB, and an effect on number of children controlling for AFLB. Moreover, these
channels often push in opposite directions.

```{r plot-age-flb, fig.cap = "Effect sizes on number of children, controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}

res_combined %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(2)),
        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
      ) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
        color = p.value < 0.05/{{n_regs}})) + 
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
        pval_geom_point() +
        pval_scale() +
        scale_fill_brewer(type = "qual", palette = 3) +
        my_labs() +
        my_theme()
```

## Weighted regressions

@@

These results show that there may be ascertainment bias in estimates of natural
selection derived from UK Biobank. One way to compensate for this is by survey
weighting. Weighting is not guaranteed to lead to unbiased estimates, because as
well as ascertainment on characteristics which we measure and weight for, there
may also be ascertainment on other, unmeasured characteristics. But the exercise
will give us a rough sense of the direction and magnitude of bias. We weight
participants using data from the 2011 Census and from the 2006 General Household
Survey (GHS). We try three alternative weighting schemes:

* Geographical weighting by census MSOA, age and presence/absence of a partner;
* Weighting by age and highest educational qualification;
* Weighting by age, highest qualification, and age at first live birth, using
  GHS data (women only).

Figure \@ref(fig:plot-res-wt) plots effect sizes among the entire sample, using
these weighting schemes. We include unweighted regressions for comparison. Table
\@ref(tab:tbl-res-wt) gives effect sizes as a proportion of the unweighted
effect size. As expected, weighting generally increases estimated effect sizes.
Sizes are reasonably similar across the geographical and age/qualification
weights, but they are increased further when we weight by AFLB. This suggests
that estimates might be further affected by weighting on other demographic
variables. Given that the Biobank sample seems to be ascertained in ways that
reduce estimates of natural selection, we suspect that these weighted estimates
are still conservative.

## Change over time

The UK Biobank data contains information on respondents' number of siblings.
Since respondents' polygenic scores are equal in expectation to their parents' 
mean scores, we can use this to look at natural selection in the parents'
generation. The dependent variable is now *number of siblings* (including
the respondent him- or herself).

The parents' generation has a second source of ascertainment bias, beyond the
sampling bias of the survey. Sampling parents of respondents inherently
overweights parents who have many children. For instance, parents of three
children will have, on average, three times more children as respondents in UK
Biobank than parents of one child. Parents of no children will
not be represented at all, by definition.

To compensate for this, we reweight our preferred weightings (Age/Qualification)
by the inverse of *number of siblings*. We then regress this on PGS.
Figure \@ref(fig:plot-res-sibs) shows the results. For comparison, we show
regressions on *number of children*, where we have excluded respondents with no 
children.

Effect sizes appear larger in the parents' generation, i.e.
for *number of siblings*. However, we treat this result cautiously, because
when we split respondents up by year of birth, we find few differences in
effect sizes between early- and late-born respondents, whether our 
dependent variable is *number of siblings* or *number of children*. In other words,
since estimated effect sizes change when we change the dependent variable,
but do not change over time within either dependent variable, the explanation
may be due to remaining ascertainment bias within our sample.

```{r plot-res-sibs, fig.cap = "Effects of polygenic scores on fertility, respondents' parents vs. respondents"}
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_children_comparison)

## Mechanisms behind modern natural selection

Why do selection effects have *opposite* signs across older and younger mothers at first live birth, and opposite signs when AFLB is controlled for? A possible answer is given by the
economic theory of fertility [@becker1960economic]. According to this,
increases in potential earned income affect fertility via two opposing channels.
There is an "income effect" by which children become more affordable, like any
other good. Since childrearing has a cost in time, there is also a substitution
effect: the opportunity cost of childrearing increases if one's market wage is
higher. The income effect would lead higher earners to have more children. The
substitution effect would lead higher earners to have fewer children. It is
often assumed that the substitution effect will dominate for lone parents, since
they have less opportunity to share childcare responsibilities, while the income
effect will dominate for couples who are able to reap the gains from
specialization.

@@

This explanation predicts that the strength of polygenic scores' effects on
fertility will correlate with the strength of their relationship with earnings.
To test this, we calculate the correlation of each PGS with estimated
earnings from respondent's first job[^est-income], and with
 education level, a phenotype which predicts earnings. We
then relate these measurements to effect sizes on number of children.
There is a clear negative relationship between a score's effect on
fertility, and its correlation with earnings or education (Figure \@ref(fig:plot-cors-earnings-educ)). If we control for age at first live birth,
the relationship changes sign and becomes positive.

[^est-income]: Earnings are estimated from the 2007 Annual Survey of Hours and
Earnings, using the SOC 2000 job code.

@@

Natural selection is taking place in modern populations on variants underlying
many polygenic scores. In a loose sense, phenotypes relating to scores which are
selected for (against) would be seen as undesirable (desirable) by many people.
For example, most people would probably prefer to have high educational
attainment, a low risk of major depressive disorder, and a low risk of coronary
artery disease, but PGS underlying these traits are being selected for. In addition,
the size of selection on PGS is negatively correlated with the score's correlation 
with household income. Potentially, natural selection could increase the health 
burden on modern societies, but that depends on effect sizes. Existing work 
argues that these are small [@beauchamp2016genetic].

However, as we show, estimates can be crucially affected by ascertainment
bias in population samples. This problem may be less serious in other surveys,
such as the US Health and Retirement Study, which explicitly aims to be
representative (as the UK Biobank does not). Nevertheless, ascertainment bias 
can still be a problem, since not all respondents consent to the collection of
genetic data: for instance, completion rates were around 80-85% for HRS genotype
data [@hrs2020].

In addition, we do not know whether estimated effect sizes of natural selection
will increase or decrease as more accurate polygenic scores are produced. We are
also unsure about whether polygenic scores for new phenotypes will show a
similar pattern of natural selection as existing ones. In short, it is probably
too early to tell whether modern natural selection is substantively important or
not.

Previous work has documented natural selection in modern populations
[@beauchamp2016genetic; @kong2017selection; @sanjak2018evidence] but has not
explained its underlying causes. The economic theory of fertility provides a
parsimonious explanation for which traits are selected, and for why the
direction of selection is reversed among certain subgroups. An alternative
theory is that traits selected for are linked to externalizing behaviour and low
time discount rates, via the channel of early sexual behaviour
[@Mills2020.05.06.081273]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and age at
smoking initiation, are selected for. However, this theory is less
good at explaining variation in selection across the full range of scores,
including for example waist-hip ratio and BMI. Externalizing behaviour also does 
not explain the reversed sign of selection among older parents. Overall, 
we believe that the economic theory is the leading explanation at present.

# Appendix

## Controlling for principal components

@@

## Selection over time

@@

>>>>

---
title: "Natural Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "`r Sys.Date()`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. The strength of natural selection on a polygenic score can be predicted by its correlation with earnings and education. Effects are larger among low-income and less educated individuals, younger parents, and people not living with a partner. Effect sizes are larger when we correct for ascertainment bias. The direction of natural selection is reversed among older parents, or  controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings can either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
---

@@

As yet we know little about the *mechanisms* behind natural selection in
contemporary populations. This paper uses data from UK Biobank to learn
more. We test for natural selection on 33 different polygenic scores. We
extend the analysis over two generations, using data on respondents'
number of siblings as well as their number of children to reconstruct
two generations of family size. This is interesting because consistent
natural selection over multiple generations could lead to substantive
effects in the long run. Most importantly, we examine the strength of
natural selection in different subgroups of the population, in order to
uncover patterns that can help illuminate the mechanisms involved.

@@

@@

We use data from UK Biobank. We compute 33 polygenic scores, normalized
to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
polygenic scores by 5-year birth intervals. Several scores show
consistent increases or declines over this 30-year period, of the order
of 5% of a standard deviation.

@@

We next regress respondents'polygenic scores on their number of children
($y_i$):

```{=tex}
\begin{equation}
y_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
\label{eq:regression}
\end{equation}
```

@@

The "selection effect", $\beta$, measures the strength of natural
selection within the sample. In fact, since polygenic scores are
normalized to mean 0, $\beta$ is the expected polygenic score among
children of the sample [@beauchamp2016genetic]. To learn more about the
underlying mechanisms, we split the sample, starting with basic
demographic variables including education, income and sex. These are all
potential sources of ascertainment bias: as well as the ascertainment
for income and education, mentioned above, the sample sex ratio skews
`r female_ratio * 100`% female.

Figure \@ref(fig:plot-income-educ-level) plots selection effects for
each polygenic score, grouping respondents by age of completing
full-time education, and by household income. Effects are larger and
more significant for the lowest income category, and for the lowest
education category. Note that the overall effect is not a simple average
of the effect among the different subgroups, because polygenic scores
may also shift respondents between the subgroups. For example, a high
PGS for educational attainment may correlate with fewer children among
early school leavers, but may also increase the age at which a
respondent leaves school.

```{r plot-income-educ-level, fig.cap = "Selection effects by education and income", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}

@@

@@

Selection effects are also different between men and women (Appendix
Figure \@ref(fig:plot-sexes)). Differences are particularly large for
educational attainment, height and MDD. Several PGS for mental illness
and personality traits are more selected for (or less against) among
women, including major depressive disorder (MDD), schizophrenia and
neuroticism, while extraversion is more selected for among men. Waist
circumference and waist-hip ratio are less selected for among women, and
educational attainment is more selected against, though the difference
is only significant for EA3.

@@

We next focus on variables related to mate search and household
formation. We split males and females by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
For both sexes, selection effects are larger and more significant among
those with more than 3 partners. Results are unchanged if we control for
respondents' age (appendix Figure
\@ref(fig:plot-n-partners-controlled)). Splitting respondents by whether
they were living with a spouse or partner at the time of interview,
effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
respondents by age at first live birth (AFLB).[^1] There is evidence for
genetic effects on AFLB [@Barban_2016], and there is a close link
between this variable and number of children born. Figure
\@ref(fig:plot-n-partners)c shows effect sizes estimated separately for
each tercile of AFLB. Several effects are strikingly different across
terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but
is not significantly selected among the youngest third. Similarly,
several PGS of body measurements are selected against only among older
mothers. The correlation between effect sizes for the youngest and
oldest terciles is `r cor_age_flb`. As before, recall that PGS may have
two effects: changing the number of children within categories, and
shifting respondents between categories.

[^1]: This information is unavailable for men.

```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Number of partners", "Household type", "Age at first live birth"), fig.ncol = 1, out.height = c("35%", "27%", "27%"), fig.align = "center"}

res_with_partner %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner"),
        with_p    = as.character(grepl("TRUE", term)),
        Household = sprintf("%s (N = %s)", Household, 
                      prettyNum(hhtype_n[with_p], big.mark = ","))
      ) %>% 
      standard_ggplot(fill = Household, n_regs = n_regs, score_col = score_name)

@@

res_age_flb_cross %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(1)),
        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
        term       = add_n(term, "age_flb_cat")
      ) %>% 
      ggplot(aes(estimate, score_name, fill = term, 
        colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 3, direction = -1, 
          guide = guide_legend(title = "Age at first live birth")) +
        my_labs() +
        my_theme()
```

@@

To investigate this further, we regress number of children on PGS
*controlling* for AFLB (appendix Figure \@ref(fig:plot-age-flb)). In
`r opp_signed` out of `r length(score_names)` cases, effects
change sign when controls are added. The correlation between effect
sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
Selection effects appear to come through two opposing channels: an
effect on AFLB, and an opposite-signed effect on number of children 
controlling for AFLB.

@@

mean_rel <- c(rel_sizes_summary[1, 1:3])
```

Since UK Biobank subjects are not representative of the wider population, our
results suggest that naïve estimates of natural selection are at risk of
ascertainment bias. To correct for this, we weight participants using data from
the 2011 Census and from the 2006 General Household Survey (GHS). We try three
alternative weighting schemes: geographical weighting by census MSOA, age and
presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Mean effect sizes across all polygenic scores are increased by 
`r mean_rel$Geographical` (geographical weighting), 
`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
further affected by weighting on other demographic variables. Since the Biobank
sample seems to be ascertained in ways that reduce estimates of selection
effects, we suspect that our weighted estimates remain conservative.

The UK Biobank data contains information on respondents' number of
siblings. Since respondents' polygenic scores are equal in expectation
to the mean scores of their parents, we can use this to look at selection 
effects in the parents' generation. We estimate equation (\ref{eq:regression})
using *number of siblings* (including the respondent) as the dependent variable.

The parents' generation has a further source of ascertainment bias: sampling
parents of respondents overweights parents who have many children. For instance,
parents of three children will have, on average, three times more children
represented in UK Biobank than parents of one child. Parents of no children will
by definition not be represented. To compensate, we reweight our preferred
weightings (Age/Qualification) by the inverse of *number of siblings*. We then
regress this on our polygenic scores. Selection effects are highly correlated across
the two generations and most share the same sign (appendix Figure \@ref(fig:plot-res-sibs)). Effect sizes appear larger in the parents' generation. 
We treat this result cautiously, because when we
split respondents up by year of birth, we find few differences in effect
sizes between early- and late-born respondents, for either generation. In other
words, since estimated effect sizes change when we change the dependent
variable, but do not change over time within either dependent variable,
the explanation may be due to remaining ascertainment bias within our
sample.

Why do selection effects have *opposite* signs across older and younger
mothers at first live birth, and opposite signs when AFLB is controlled
for? A possible answer is given by the economic theory of fertility
[@becker1960economic]. According to this, increases in potential earned
income affect fertility via two opposing channels. There is an "income
effect" by which children become more affordable, like any other good.
Since childrearing has a cost in time, there is also a "substitution
effect": the opportunity cost of childrearing increases if one's market wage is
higher. The income effect would lead higher earners to have more children. The
substitution effect would lead higher earners to have fewer children. It is
often assumed that the substitution effect will dominate for lone parents, or
those in unstable relationships, since they have less opportunity to share
childcare responsibilities, while the income effect will dominate for couples
who are able to reap the gains from specialization.

@@

This explanation predicts that the strength of polygenic scores' effects
on fertility will correlate with the strength of their relationship with
earnings. To test this, we calculate the correlation of each PGS with
estimated earnings from respondent's first job[^earnings-est], and with education
level, a phenotype which predicts earnings. We then relate these
measurements to effect sizes on number of children. There is a clear
negative relationship between a score's effect on fertility, and its
correlation with earnings or education. If we control for age at first live
birth, the relationship changes sign and becomes positive  (Figure
\@ref(fig:plot-cors-earnings-educ)).

[^earnings-est]: Earnings are estimated from the 2007 Annual Survey of Hours and
    Earnings, using the SOC 2000 job code.

@@

Previous work has documented natural selection in modern populations on variants
underlying many polygenic scores [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that natural selection effects are highly
concentrated among specific subgroups of the population, including poorer
people, less educated people, younger parents and those with more lifetime
sexual partners. Indeed, among older parents selection effects are actually
reversed. Furthermore, the size of selection effects correlates with the
association between a PGS and labour market earnings. The economic theory of
fertility provides a parsimonious explanation for this.

An alternative theory is that traits selected for are linked to
externalizing behaviour and low time discount rates, via the channel of
early sexual behaviour [@Mills2020.05.06.081273]. The data here provide
some support for this: scores which might plausibly be linked to
externalizing behaviour, like ADHD and age at smoking initiation, are
selected for. However, this theory is less good at explaining variation
in selection across the full range of scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing behaviour also does not explain
that selection effects change sign among older parents. Overall, we believe
that the economic theory is the leading explanation at present.

PGS which correlate with high (low) earnings and more (less) education are being
selected against (selected for). In addition, many of the phenotypes under
positive selection are linked to poor health, or would be seen as undesirable to
have by many people. For example, most people would probably prefer to have high
educational attainment, a low risk of major depressive disorder, and a low risk
of coronary artery disease, but natural selection is pushing against these
traits. Potentially, this could increase the health burden on modern societies,
but that depends on effect sizes.

Our results suggest that naïve estimates can be affected by sample ascertainment
bias. This problem may be less serious in surveys which aim to be representative
(as the UK Biobank does not). Nevertheless, not all respondents consent to the
collection of genetic data: for instance, completion rates for genotype data in
the US Health and Retirement Study  were around 80-85% [@hrs2020]. Researchers
should be aware of the risks of ascertainment when studying contemporary 
natural selection.

In addition, we do not know whether estimated effect sizes of natural
selection will increase or decrease as more accurate polygenic scores
are produced. We are also unsure whether genetic variants underlying other
phenotypes will show a similar pattern of natural selection as these ones. 
In short, it is probably too early to tell whether modern natural
selection is substantively important or not.

Research on natural selection is controversial. It brings to mind eugenics,
which has been used to justify mass murder and human rights violations, in the
past and possibly the present [@mao1998chinese; @sleeboom2011genetic;
@zenz2020]. Whilst we claim scientific legitimacy for our empirical results, we
do not claim to speak with any special authority on how society should react to
them. This can only be resolved by open debate, in which scientists take part as
equals among many. We therefore choose not to discuss policy implications here.

\FloatBarrier
# Appendix

## Natural selection by sex

```{r plot-sexes, fig.cap = "Selection effects by sex. Lines show significant differences."}

\FloatBarrier
## Controlling for AFLB

```{r plot-age-flb, fig.cap = "Selection effects controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}

res_combined %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(2)),
        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for AFLB")
      ) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
        color = p.value < 0.05/{{n_regs}})) + 
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
        pval_geom_point() +
        pval_scale() +
        scale_fill_brewer(type = "qual", palette = 3) +
        my_labs() +
        my_theme()
```

\FloatBarrier
## Weighted regressions

Figure \@ref(fig:plot-res-wt) plots effect sizes among the entire
sample, using our three weighting schemes. We include unweighted regressions
for comparison. Table \@ref(tab:tbl-res-wt) gives effect sizes as a
proportion of the unweighted effect size.

\FloatBarrier
## Parents' generation

Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
polygenic scores. For comparison, we show regressions on the respondents'
generation (*number of children*), excluding respondents with no
children.

```{r plot-res-sibs, fig.cap = "Selection effects, respondents' parents vs. respondents"}
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_children_comparison)

\FloatBarrier
## Controlling for principal components

@@

\FloatBarrier
## Selection over time

@@

\FloatBarrier
